---
title: "Analysis Assignment 3"
author: "Stat 847"
date: "Due: Thursday March 17, Name: Gurmanjot Singh, Waterloo ID: 21217394"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(crop = NULL)
```

# Preamble

Your assignment must be submitted by the due date listed at the top of this document, and it must be submitted electronically in .pdf format via Crowdmark.

This means that your responses for different question parts should begin on separate pages of your .pdf file. Note that your .pdf solution file must have been generated by R Markdown. Additionally:

Organization and comprehensibility is part of a full solution. Consequently, points will be deducted for solutions that are not organized and incomprehensible. Furthermore, if you submit your assignment to Crowdmark, but you do so incorrectly in any way (e.g., you upload your Question 2 solution in the Question 1 box), you will receive a 5% deduction (i.e., 5% of the assignment’s point total will be deducted from your point total).

There are 27 marks in total.

\newpage

# Dataset

# Use the IMDB API

All the data for this assignment is to be downloaded from IMDB (the Internet Movie Database) and formatted using `fromJSON` in the `jsonlite` package in R (or equivalent in Python, or Julia).

To use the internet IMDB API, consult the guide here: <https://imdbapi.dev/> and pull the request URLS. This API does not require a token unless we are doing more than 1000 API queries in a day (and this assignment is fewer than 300 queries).

To see how to get usable data from such a request URL, look at "API in eSports", especially the section that starts on Slide 6 when `fromJSON` in the `jsonlite` package is discussed.

```{r}
# Imports for assignment
library(jsonlite)
library(httr)
library(dplyr)
library(ggplot2)
```

\newpage

# Question 1 (2 marks):

Using the IMDB API, get the `originalTitles` from the first five movies (make sure to specific type as movie) made in 2025 (both `startYear` and `endYear` should be 2025) that have a vote count of at least 10000. Hint: The first one on your list should be "Marty Supreme".

```{r}
# API call for first 50 movies with at least 10000 votes
url_q1 <- "https://api.imdbapi.dev/titles?types=MOVIE&startYear=2025&endYear=2025&minVoteCount=10000"

response <- GET(url_q1, config(ssl_verifypeer = FALSE))

# Get first page from JSON
page1 <- fromJSON(content(response, as = "text", encoding = "UTF-8"))


# First five movies
page1$titles$originalTitle[1:5]
```

The first one is "Marty Supreme" as expected.

\newpage

# Question 2 (2 marks):

Using the `$nextPageToken` that was given as part of your API response in the last question, get the `originalTitles` of the 51st to 55th movies made in 2025 with at least 10000 votes. (Hint: The 51st one should be "The Naked Gun")

If you're looking for `$nextPageToken`, use `str` to see the structure and first few elements in everything in the JSON object you pull.

```{r}
# Get the next page token
next_token <- page1$nextPageToken

# 
url_q2 <- paste0(
  "https://api.imdbapi.dev/titles?types=MOVIE&startYear=2025&endYear=2025",
  "&minVoteCount=10000&pageToken=", next_token
)

response2 <- GET(url_q2, config(ssl_verifypeer = FALSE))

# 
page2 <- fromJSON(content(response2, as = "text", encoding = "UTF-8"))
page2$titles$originalTitle[1:5]
```

The first one is "The Naked Gun" as expected.

\newpage

# Question 3 (6 marks):

Extract a dataset of all movies with at least 10000 votes with `startYear` 1976 and `endYear` and 2025 inclusive. Give the number of such movies. The final dataset should have about 11,000 rows (you can find the exact number in the structure of the JSON object of the first page), which means you'll need to loop about 220 times.

Report the first and last 6 rows of the dataset with `head()` and `tail()` or similar, **and show your completed code**.

**HINT 1:** Use a framework like the one below to make sure that you don't get blocked out of the API (specifically, one that waits a few seconds between each API call)

```{r, eval=FALSE}

# Helper for shortening getting pages
fetch_page <- function(url) {
  fromJSON(content(GET(url, config(ssl_verifypeer = FALSE)), as = "text", encoding = "UTF-8"))
}

# First page API call
base_url <- paste0(
  "https://api.imdbapi.dev/titles?types=MOVIE",
  "&startYear=1976&endYear=2025",
  "&minVoteCount=10000"
)

first_page <- fetch_page(base_url)

# Get total number of movies for loop
totalCount <- first_page$totalCount

# Wrapping titles as data frame and setting rownames to avoid duplicates
df_movie <- data.frame(first_page$titles)
rownames(df_movie) <- 1:50

# Page traversal loop
count <- 50
this_page <- first_page

while (count < totalCount) {
  next_token <- this_page$nextPageToken
  url <- paste0(base_url, "&pageToken=", next_token)
  
  this_page <- fetch_page(url)
  new_rows <- data.frame(this_page$titles)
  rownames(new_rows) <- (count + 1):(count + nrow(new_rows))
  
  df_movie <- bind_rows(df_movie, new_rows)
  count <- count + nrow(new_rows)
  
  print(paste("Fetched:", count, "/", totalCount))
  Sys.sleep(5)
}

# Clean nested columns
df_movie$primaryImage    <- NULL
df_movie$aggregateRating <- df_movie$rating$aggregateRating
df_movie$voteCount       <- df_movie$rating$voteCount
df_movie$rating          <- NULL
df_movie$plot            <- as.character(df_movie$plot)
df_movie$genres          <- as.character(df_movie$genres)
df_movie                 <- data.frame(df_movie)

write.csv(df_movie, "IMDB_Major_Titles_1976-2025.csv", row.names = FALSE)
```

**HINT 2:** Do this only once and save the result as a CSV file or similar so that you can retrieve it later without calling the API many times and waiting. IF you're showing your code in Rmarkdown use `r, eval=FALSE` for the API call, and `r` for loading it from a CSV file. That way it won't take forever to make your R markdown file.

```{r}
# Reading csv
df_movie <- read.csv("IMDB_Major_Titles_1976-2025.csv")

# Show head and tail and number of rows
head(df_movie)
tail(df_movie)
nrow(df_movie)
```

\newpage

# Question 4 (2 marks):

\*\*If you are stuck on Questions 1-3, you can use the dataset `IMDB_Major_Titles_1976-2025.csv` instead of your own.

Make a side-by-side boxplot where every box represents the middle half of the runtime lengths of movies with that `startYear`. Comment on the trend you see in movie lengths over the last 50 years. Set the y limits to 1-5 hours (3600-12000 seconds), and use an appropriate title and axis labels.

```{r}
# Side-by-side boxplot of runtime by year
ggplot(df_movie, aes(x = factor(startYear), y = runtimeSeconds)) +
  geom_boxplot() +
  scale_y_continuous(
    limits = c(3600, 12000),
    breaks = seq(3600, 12000, by = 1800),
    labels = c("1h", "1.5h", "2h", "2.5h", "3h")
  ) +
  labs(
    title = "Distribution of Movie Runtimes by Year (1976-2025)",
    x = "Year",
    y = "Runtime"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))
```

The median movie length appears to remain relatively stable just below 2 hours, while the variation in runtimes appears to get a little larger. Most outliers stay above most runtimes in the 2.25-3.25 hour range, and get denser over time, likely due to more releases per year.

\newpage

# Question 5 (2 marks):

Draw a line plot of the average length of movie by year, using the bounds 1.5-2 hours (5400 to 7200 seconds). Does your description from question 4 hold?

```{r}
# Line plot of average runtime by year
avg_runtime <- aggregate(runtimeSeconds ~ startYear, data = df_movie, FUN = mean, na.rm = TRUE)

ggplot(avg_runtime, aes(x = startYear, y = runtimeSeconds)) +
  geom_line() +
  geom_point(size = 1.5) +
  scale_y_continuous(
    limits = c(5400, 7200),
    breaks = seq(5400, 7200, by = 450),
    labels = c("1.5h", "1.625h", "1.75h", "1.875h", "2h")
  ) +
  labs(
    title = "Average Movie Runtime by Year (1976-2025)",
    x = "Year",
    y = "Average Runtime"
  ) +
  theme_minimal()
```

The average runtime appears to be heavily influenced by the outliers, considering how its shape is very close to the vague movement of the outlier graph. This verifies that the main body of the points don't contribute a visible trend to the average, and so are relatively consistent just below 2 hours, which is reflected in the range the average resides in.

\newpage

# Question 6 (4 marks):

For each year from 1976-2025, find the movie that has the most votes, and make a dataset of them. Sort from earliest year to latest, and report the first six and last six rows of this dataset (makes sure to keep the `id`, `aggregateRating`, and `voteCount` of each movie, as well as the `originalTitle`.) You can assume there are no ties for the most votes.

```{r}
# Ensure distinct movies
df_movie <- df_movie %>% distinct(id, .keep_all = TRUE)

# Most-voted movie per year
most_voted <- df_movie %>%
  group_by(startYear) %>%
  slice_max(voteCount, n = 1) %>%
  ungroup() %>%
  arrange(startYear) %>%
  select(startYear, id, originalTitle, aggregateRating, voteCount)

# Head and tail
head(most_voted)
tail(most_voted)
nrow(most_voted)  # should be exactly 50
```

\newpage

# Question 7 (4 marks):

use the API to get the box office information for the 50 titles from Question 6 by using the `id` values (e.g., `tt0075314` for Taxi Driver). From each box office info, get the `$amount` for the `domesticGross`, `worldwideGross`, and `productionBudget`. Append all these onto your dataset from question 6 as new variables.

Report the first and last six rows of the data set with the three new variables.

```{r, eval=FALSE}
# Helper for getting box office info
fetch_boxoffice <- function(id) {
  url <- paste0("https://api.imdbapi.dev/titles/", id, "/boxOffice")
  r <- GET(url, config(ssl_verifypeer = FALSE))
  if (status_code(r) == 200) {
    p <- fromJSON(content(r, as = "text", encoding = "UTF-8"))
    return(data.frame(
      id = id,
      domesticGross    = as.numeric(p$domesticGross$amount),
      worldwideGross   = as.numeric(p$worldwideGross$amount),
      productionBudget = as.numeric(p$productionBudget$amount)
    ))
  } else {
    return(data.frame(id = id, domesticGross = NA, worldwideGross = NA, productionBudget = NA))
  }
}

# 
boxoffice_list <- lapply(most_voted$id, function(id) {
  result <- fetch_boxoffice(id)
  Sys.sleep(1)
  result
})

boxoffice_df <- bind_rows(boxoffice_list)

# Merge onto most_voted
most_voted <- left_join(most_voted, boxoffice_df, by = "id")

# For knitting purposes
write.csv(most_voted, "most_voted_boxoffice.csv", row.names = FALSE)
```

```{r}
# Read to avoid calling API
most_voted <- read.csv("most_voted_boxoffice.csv")

# Updated head and tail
head(most_voted)
tail(most_voted)
```

\newpage

# Question 8 (2 marks):

Plot the aggregateRating and the domesticGross together. Explain any trend you see.

```{r}
# aggregateRating vs domesticGross
ggplot(most_voted, aes(x = aggregateRating, y = domesticGross)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "IMDb Rating vs Domestic Gross for Most-Voted Movie per Year (1976-2025)",
    x = "IMDb Rating",
    y = "Domestic Gross (USD)"
  ) +
  theme_minimal()
```

There is a slight average upward trend but the points don't appear to cluster well around it or in another clear way, so the most voted movies don't appear to have their earnings and ratings strongly correlated. However, we do see that for ratings below 8.0, there are only three points, while for higher ratings there are many more movies that are highly voted on, particularly in the 8-9 rate range. This bias is likely due to movies people like being more popular on average, so they are rated pretty well and so are found primarily in this range. The variance on the earnings is also higher in this range, with generally larger maximums even after ignoring the two extremely high outliers. Notably we also don't have many points past 9.0, which is likely due to extremely high voting ensuring the average is more likely to stay at something people consider as a reasonable good rating, rather than too embellishing.

\newpage

# Question 9 (3 marks):

Make a model of worldwideGross as function of domesticGross, and comment on the three movies with the largest positive residuals (worldwideGross is large compared to domesticGross), and the largest negative residuals.

```{r}
# worldwideGross ~ domesticGross
model <- lm(worldwideGross ~ domesticGross, data = most_voted)
summary(model)

most_voted$residuals <- residuals(model)

# 3 largest positive residuals (worldwide >> domestic)
most_voted %>% arrange(desc(residuals)) %>% 
  select(startYear, originalTitle, domesticGross, worldwideGross, residuals) %>% 
  head(3)

# 3 largest negative residuals (worldwide << domestic)
most_voted %>% arrange(residuals) %>% 
  select(startYear, originalTitle, domesticGross, worldwideGross, residuals) %>% 
  head(3)
```

The three most positive residuals are: "Titanic", "Harry Potter and the Deathly Hallows: Part 2", and "Interstellar". This could be because of the wide international appeal of the stories, the first being a movie about a historical event almost everyone knows about as a romance, the second connected to very popular books all over the world, and the third is a sci-fi epic with general appeal as a story. It could also be because of the way it was released, all three had international releases and were very publicized and so many different audiences could access them.

The three most negative residuals are: "Star Wars", "The Dark Knight", and "Sinners". The first two are possibly due to how Star Wars and DC movies are embedded in their domestic (American) pop culture, that didn't translate proportionally to international audiences at their times. For the third, Sinners was a story specifically set in American South cultural background and Black history, so it could have resonated better with an American audience.

# Footnotes

Relevant articles for your interest

<https://movies.stackexchange.com/questions/33815/when-discussing-box-office-revenue-why-do-we-typically-only-refer-to-the-domesti>

<https://www.mangosorbananas.com/p/the-declining-appeal-of-hollywood>

<https://www.t3mediastudios.com/post/movie-budgets-exposed-the-real-cost-of-a-blockbuster-then-vs-now>
