---
title: "STAT 847 W26 Analysis Assignment 2"
subtitle: "Due: Monday February 23, Name: Gurmanjot Singh, Waterloo ID: 21217394"
output:
  pdf_document:
    keep_tex: yes
    number_sections: no
  html_document:
    toc: yes
  word_document: default
urlcolor: blue
---

# Preamble

Your assignment must be submitted by the due date listed at the top of this document, and it must be submitted electronically in .pdf format via Crowdmark.

This means that your responses for different question parts should begin on separate pages of your .pdf file. Note that your .pdf solution file must have been generated by R Markdown. Additionally:

Organization and comprehensibility is part of a full solution. Consequently, points will be deducted for solutions that are not organized and incomprehensible. Furthermore, if you submit your assignment to Crowdmark, but you do so incorrectly in any way (e.g., you upload your Question 2 solution in the Question 1 box), you will receive a 5% deduction (i.e., 5% of the assignment’s point total will be deducted from your point total).

There are 30 marks in total.

\newpage

# Dataset

This dataset is from an ongoing research project called the Dragon Hoard Index looking at the valuation history of collectibles. It has the following variables. There are 4992 measurements across 38 different coins with 3 different grades each.

Your general goal with this dataset is to make various predictive models of the coin prices, and ultimately an index to track the Canadian coin market as a whole.

| Variable    | Explanation                                            |
|-------------|--------------------------------------------------------|
| url         | URL where the pricing information was extracted from   |
| coin        | Code for the denomination and minting year of the coin |
| grade       | Condition of the coin, VF-20 (Very fine), ...          |
| ...         | ...EF-40 (Extremely fine), and MS-60 (Mint state)      |
| price_year  | Year that the collector price of the coin was measured |
| price       | Collector price of the coin                            |
| circulation | Number of copies of the coin that were minted          |
| denom_cents | Denomination in cents, from 1 to 100                   |
| mint_year   | Year that the coin that minted                         |
| age         | Age in years of the coin when the price was measured   |
| pf_ratio    | Ratio of Price to Face value.                          |

```{r}
# Imports for rest of assignment
library(glmnet)
library(randomForest)
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(plyr)
library(dplyr)
```

```{r}
df_coin = read.csv("Canada Coin Price History 2026-02-09.csv")
head(df_coin)
dim(df_coin)
length(unique(df_coin$coin))
```

\newpage

# Questions

## Q1) (4 marks)

Make some derivative variables to make things more convenient for later analysis. Specifically.

-   Make `denom_cents` a categorical variable, call it `denomination`
-   Make grade an ordinal variable `VF-20 < EF-40 < MS-60`, call it `grade_ordinal` (Use the lookup `?factor` for details)
-   Make a $10^6/x$ transformation of circulation, call it `rarity`.
-   Make a natural log of price, call it `log_price`

Show the 1st, 1001th, 2001st, 3001st, and 4001st row of the data with the new variables.

```{r}
# Make denom_cents categorical
df_coin$denomination <- as.factor(df_coin$denom_cents)

# Make grade ordinal
df_coin$grade_ordinal <- factor(df_coin$grade, 
                                levels = c("VF-20", "EF-40", "MS-60"),
                                ordered = TRUE)

# Transform circulation to rarity
df_coin$rarity <- 10^6 / df_coin$circulation

# Log of price
df_coin$log_price <- log(df_coin$price)

# Relevant rows
rows_to_show <- c(1, 1001, 2001, 3001, 4001)
df_coin[rows_to_show, ]
```

\newpage

## Q2) (2 marks)

Make a `lm()` linear model of the `log_price` of coins as a function of

-   price_year
-   age
-   grade_ordinal
-   denomination
-   rarity

...without any complications like transformations and interactions.

Name the model `modq2`, and report the `summary` of it.

```{r}
# Model
modq2 <- lm(log_price ~ price_year + age + grade_ordinal + denomination + rarity,
            data = df_coin)

# Summary
summary(modq2)
```

\newpage

## Q3) (4 marks)

Starting with the `modq2` variables, make a simpler model with LASSO. For the penalty term, find and report the best lambda value as defined by `cv.glmnet`. Give a short (50 words or fewer) write-up of which variables made it into the final model and which did not. Make a guess as to what's important and why.

```{r}
# Prepare data for glmnet
x <- model.matrix(log_price ~ price_year + age + grade_ordinal + denomination + 
                  rarity,data = df_coin)[, -1]
y <- df_coin$log_price

# Find best lambda
set.seed(123)
cv_lasso <- cv.glmnet(x, y, alpha = 1)
best_lambda <- cv_lasso$lambda.min
cat("Best lambda:", best_lambda, "\n")

# Fit LASSO
modq3 <- glmnet(x, y, alpha = 1, lambda = best_lambda)

# Coefficients
coef(modq3)
```

No variables were cut since the best lambda was small. Largest positive contribution came from grade_ordinal.L (positive linearity in grades) and denominations (especially 20 and 50) besides 100. Rarity and age also have small positive effect on price, although smaller than the others.

\newpage

## Q4) (4 marks)

Using the `randomForest` function in `library(randomForest)`, make a random forest using the variables in `modq2`. Give your forest 1000 trees, have each tree use 5 variables, and set a minimum node size of 5. Sample with replacement. Report the percentage of variance explained by the forest using `print()`.

```{r}
# Prepare random forest data
rf_data <- df_coin[, c("log_price", "price_year", "age", "grade_ordinal", 
                       "denomination", "rarity")]
rf_data <- na.omit(rf_data)

# Build RF
set.seed(123)
modq4 <- randomForest(log_price ~ price_year + age + grade_ordinal + denomination + rarity,
                      data = rf_data,
                      ntree = 1000,
                      mtry = 5,
                      nodesize = 5,
                      replace = TRUE)

# Variance explained
print(modq4)
```

\newpage

## Q5) (3 marks)

Use `predict()` or equivalent to fit the price by taking the exponent of the predicted `log_price` from the models you made in Q2, Q3, and Q4. Report the fitted values for the 1st, 1001st, 2001st, 3001st, and 4001st rows in the dataset. Also report the true values for comparison.

For the Q3 predictions, set the newx to the `data.matrix` object that you used to make `x` for `modq3`.

```{r}
# Q2 predictions (linear model)
pred_q2_log <- predict(modq2, newdata = df_coin[rows_to_show, ])
pred_q2 <- exp(pred_q2_log)

# Q3 predictions (LASSO)
pred_q3_log <- predict(modq3, newx = x[rows_to_show, ], s = best_lambda)
pred_q3 <- exp(pred_q3_log)

# Q4 predictions (Random Forest)
pred_q4_log <- predict(modq4, newdata = df_coin[rows_to_show, ])
pred_q4 <- exp(pred_q4_log)

# Create comparison table
predictions_df <- data.frame(
  Row = rows_to_show,
  True_Price = df_coin$price[rows_to_show],
  LM_Predicted = pred_q2,
  LASSO_Predicted = as.vector(pred_q3),
  RF_Predicted = pred_q4
)
print(predictions_df)
```

Note that the LM and LASSO predictions are close to each other due to the similar parameters from not having a big lambda as mentioned before. The RF predictions are much closer to the true prices, even though all the predictions deviate more at the later rows.

\newpage

## Q6) (4 marks)

Find the best tuning parameter (in this case, the minimum node size) for the models you made in Q4 by finding the model with the lowest mean squared error for the random forest. Keep the number of variables to try at 5, and tune on the minimum node size 5, 10, 15, ... , 100. For speed purposes, only make 200 trees in each candidate forest.

You can find the mean squared error with `modq4$mse[200]`

Use the best minimum node size and make a model with 1000 trees and get the predictions of the 1st, 1001st, 2001st, 3001st, and 4001st rows in the dataset like you did in Q5 for the random forest, comment in one or two sentences whether these predictions fit the actual data better.

```{r}
# Tune minimum node size
node_sizes <- seq(5, 100, by = 5)
mse_results <- numeric(length(node_sizes))

set.seed(123)
for(i in 1:length(node_sizes)) {
  rf_temp <- randomForest(log_price ~ price_year + age + grade_ordinal + denomination + rarity,
                          data = rf_data,
                          ntree = 200,
                          mtry = 5,
                          nodesize = node_sizes[i],
                          replace = TRUE)
  mse_results[i] <- rf_temp$mse[200]
}

# Find best node size
best_nodesize <- node_sizes[which.min(mse_results)]
cat("Best minimum node size:", best_nodesize, "\n")
cat("MSE for each node size:\n")
data.frame(NodeSize = node_sizes, MSE = mse_results)

# Build final model with best node size
set.seed(123)
modq6 <- randomForest(log_price ~ price_year + age + grade_ordinal + denomination + rarity,
                      data = rf_data,
                      ntree = 1000,
                      mtry = 5,
                      nodesize = best_nodesize,
                      replace = TRUE)

# Get predictions for comparison
pred_q6_log <- predict(modq6, newdata = df_coin[rows_to_show, ])
pred_q6 <- exp(pred_q6_log)

predictions_q6 <- data.frame(
  Row = rows_to_show,
  True_Price = df_coin$price[rows_to_show],
  RF_Original = pred_q4,
  RF_Tuned = pred_q6
)
print(predictions_q6)
```

The tuned RF consistently is further from the true price than the original RF, always estimating on the same side but just a larger error margin. Note that our choice of best node size was tuned for RFs with 200 trees, a larger node size (here being 40) would be more suitable here so the smaller number of trees don't overfit to the data due to the higher variance.

In the RF with 1000 trees, the averaging reduces the variance and risk of overfitting, so we can reasonably increase the expressiveness with a smaller node size. In fact, having too large a node size can push the model towards underfitting, which is what we see in this case, where our naive selection of 40 was worse than the original RF with node size 5.

\newpage

## Q7) (5 marks)

Make a new variable `grade_numeric` which is just `as.numeric(df_coin$grade_ordinal)`

Using the `PCA()` function in the `factoMineR` package, build a principal component analysis on the following variables in `df_coin`: `grade_numeric`, `price_year`, `log_price`, `rarity`, `denom_cents`, and `age`.

Report the scree plot (uses the `factoextra` package, which needs `ggplot2`) and the variables factor map. Comment in 1-2 sentences on each plot.

This guide may be useful: <https://www.sthda.com/english/wiki/wiki.php?id_contents=7851>

```{r}
# Make grade_numeric
df_coin$grade_numeric <- as.numeric(df_coin$grade_ordinal)

# Prepare data for PCA
pca_data <- df_coin[, c("grade_numeric", "price_year", "log_price", 
                        "rarity", "denom_cents", "age")]
pca_data <- na.omit(pca_data)

# Perform PCA
pca_result <- PCA(pca_data, graph = FALSE)

# Scree plot
fviz_eig(pca_result, addlabels = TRUE)

# Variables factor maps
# PC1 vs PC2
fviz_pca_var(pca_result, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))

# PC1 vs PC3
fviz_pca_var(pca_result, axes = c(1, 3), col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))

# PC2 vs PC3
fviz_pca_var(pca_result, axes = c(2, 3), col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
```

Looking at the scree plot, the first three principal components seem to carry most of the explained variance (85% about), and it flattens out after them, so we can reduce the effective dimensionality of our data to 3 for variation along those components.

To capture the variable factors map for 3 components, we would need three snapshots of each pair of components, provided in the output.

If we compare the directions across the three pictures, it appears that price_year is strongly pointing in the PC2 direction, with age and log_price nearby but not completely aligned (some of both is in PC1 and PC3). Likewise, rarity and denom_cents have high contributions to PC1 and are highly correlated, while grade_numeric makes up for most of PC3, as it vanishes outside.

\newpage

## Q8) (5 marks)

Make a new variable called `price_index` which is the calculated by the value of the coin at a given grade divided by its 1960 price, times 100. Leave `price_index` as `NA` for any coin that does not have a price for that grade in 1960 (there might be no such coins) For example, the VF-20 grade 1 cent 1858 coin for the years 1958-1963 should have index values of 67.35, 83.67, 100, 112.25, 128.57, 145.58 (up to rounding) respectively.

```{r}
# Reference table of 1960 prices, average out duplicates
prices_1960 <- df_coin %>%
  filter(price_year == 1960) %>%
  group_by(coin, grade) %>%
  summarise(price_1960 = mean(price), .groups = 'drop')

# Join back and calculate index
df_coin <- df_coin %>%
  left_join(prices_1960, by = c("coin", "grade")) %>%
  mutate(price_index = ifelse(!is.na(price_1960), 
                              (price / price_1960) * 100, 
                              NA))

# Verify with the example (1 cent 1858 coin, VF-20 grade)
example_subset <- df_coin %>%
  filter(coin == "1-cent-1858", grade == "VF-20", price_year %in% 1958:1963) %>%
  select(coin, grade, price_year, price, price_index)

print(example_subset)
```

The example's expected indices match.

\newpage

## Q9) (6 marks)

**USING ONLY VF-20 GRADE COINS**, make a new dataset using ddply (or equivalent, including equivalents in Python or Julia), where one column is one `price_year`, and the other is `coin_index`, a weighted average of `price_index`, where the weight is `circulation`. In other words, a coin with 2 million circulation should count for twice as much in the `coin_index`.

Create a broken line graph (or equivalent in Python or Julia) of `coin_index` as a function of `price_year`.

```{r}
# Filter for VF-20 grade only
vf20_data <- df_coin[df_coin$grade == "VF-20" & !is.na(df_coin$price_index), ]

# Weighted average price_index by price_year
coin_index_df <- ddply(vf20_data, .(price_year), summarize,
                       coin_index = weighted.mean(price_index, circulation, na.rm = TRUE))

# Create line graph
ggplot(coin_index_df, aes(x = price_year, y = coin_index)) +
  geom_line() +
  geom_point() +
  labs(title = "Canadian Coin Index (VF-20 Grade)",
       x = "Year",
       y = "Coin Index (1960 = 100)") +
  theme_minimal()
```

\newpage

## Q10) (6 marks)

Predict the 1st, 1001st, 2001st, 3001st, and 4001st observations again, but do so by taking the 1960 price for the coin question in the grade in question, and applying the appropriate `coin_index` for that `price_year`. For example, for the 1001st observation, look at the MS-60 1-cent 1960 coin price. Comment on three things, one sentence each:

-   The performance of the prediction compared to the predictions in Q5 and Q6.
-   The appropriateness of this prediction across these various coin grades.
-   The convenience of being able to apply this prediction model to coins not included in the model.

I will interpret the question as using the VF-20 coin index as an estimate of the index for the other grades, since we only have that index from question 9.

```{r}
# Create predictions for specified rows
pred_q10 <- numeric(length(rows_to_show))

for(i in 1:length(rows_to_show)) {
  row_idx <- rows_to_show[i]
  
  # Get coin, grade, and price_year for this observation
  coin_id <- df_coin$coin[row_idx]
  grade_val <- df_coin$grade[row_idx]
  year_val <- df_coin$price_year[row_idx]
  
  # Get 1960 price for this coin-grade
  price_1960_idx <- which(df_coin$coin == coin_id & 
                          df_coin$grade == grade_val & 
                          df_coin$price_year == 1960)
  
  if(length(price_1960_idx) > 0) {
    price_1960 <- df_coin$price[price_1960_idx[1]]
    
    # Get coin_index for this year
    index_val <- coin_index_df$coin_index[coin_index_df$price_year == year_val]
    
    # Get index scaled prediction
    if(length(index_val) > 0) {
      pred_q10[i] <- price_1960 * (index_val / 100)
    } else {
      pred_q10[i] <- NA
    }
  } else {
    pred_q10[i] <- NA
  }
}

# Relevant predictions
predictions_q10 <- data.frame(
  Row = rows_to_show,
  Coin = df_coin$coin[rows_to_show],
  Grade = df_coin$grade[rows_to_show],
  Year = df_coin$price_year[rows_to_show],
  True_Price = df_coin$price[rows_to_show],
  Index_Predicted = pred_q10
  
)
print(predictions_q10)
```

The index prediction is much closer to the true price for the first three rows compared to the original linear model, LASSO, and the two RFs, but severely undershoots at 3001 and 4001 (much more than the aforementioned models), which could be due to the index not accounting for price variations across different grades (as well as averaging over coin variations), especially at later years well beyond 1960.

Because it does not account for this variation over grades, it is not really appropriate for cross-grade prediction, but it does appear to have some predictive power for the coins with lower denominations regardless of the grade.

This method is generally convenient for any (Canadian) coin with an available 1960 price, since we can just scale it using the index, although we have no guarantees for how reasonable or useful such a model would be depending on the history of that coin, especially in recent years.
